\chapter{Literature Review}
\label{ch:litreview}

\section{Critical Assessment of Prior Works}
We synthesise and critically analyse the current state of student simulation and Socratic tutoring, highlighting achievements, trajectories, omissions, and how our work addresses these gaps.

\subsection{Epistemic Fidelity vs. Surface Realism}
Yuan et al. \cite{yuan2026towards} established the concept of Epistemic State Specification (ESS) for valid student simulation, arguing that simulators must prioritise the accuracy of the learner's knowledge and reasoning processes over mere dialogue fluency. However, their framework is primarily conceptual and assumes static learner capabilities. Scarlatos et al. \cite{scarlatos2026simulated} benchmarked various simulators, finding that standard prompting strategies suffer heavily from sycophancy bias and fail to maintain consistent misconception-driven belief states. To standardise evaluation across these dimensions, Maurya et al. \cite{maurya2025unifying} propose a unified pedagogical evaluation taxonomy consisting of eight dimensions (including mistake identification, mistake location, and revealing of the answer) and introduce the MRBench math tutoring benchmark. While MRBench provides a rigorous protocol for static dialogue evaluation, it is limited to mathematical domains and does not evaluate adaptive tutoring policies in interactive programming sandboxes.

\subsection{Tutor-Student Agent Interactions}
Frameworks like PETITE \cite{ozdemir2026enhancing} use asymmetric role assignment (a coder student and a helper tutor) to improve code generation success, but they focus purely on task completion rather than pedagogical simulation, lacking explicit models of human cognitive fatigue. SocraticLM \cite{liu2024socraticlm} fine-tunes a model on a Socratic dialogue dataset (SocraTeach) using a Dean-Teacher-Student pipeline, but the tutor agent remains vulnerable to answer leakage when students show persistent confusion. Similarly, the KELE framework \cite{peng2025kele} introduces a structured Socratic teaching methodology via a collaborative `Consultant-Teacher' multi-agent mechanism. While KELE structures dialogue stages using a rule system (SocRule) and demonstrates Socratic performance improvements by fine-tuning on a science-themed dialogue dataset (SocratDataset), its planning is guided by qualitative LLM heuristics. It lacks explicit mathematical formulations of the student's cognitive state dynamics (such as fatigue and friction) and does not mitigate the sycophancy bias of its student simulator. To control dialogue alignment more precisely, Petukhova \& Kochmar \cite{petukhova2025intent} expand Socratic taxonomies (such as MathDial) to 11 fine-grained pedagogical intents, demonstrating that training tutors on specific pedagogical goals improves response alignment. However, their intent-driven mapping is static and does not dynamically adjust to estimated student knowledge states or active misconceptions.

\subsection{Behavioural Simulation \& Persona Modelling}
Cunling Bian \cite{bian2026generative} utilises structured student interviews and multi-expert reflection to simulate learning behaviours (measured by the MSLQ questionnaire) with high accuracy, but does not deploy agents in a live, interactive dialogue environment. Classroom Simulacra \cite{xu2025classroom} incorporates online slide interactions and webcam-driven eye-tracking (GazeRecorder) to capture student engagement, but focuses on classroom-level dynamics rather than dyadic tutor policy optimisation. GPTeach \cite{markel2023gpteach} builds interactive TA training sandboxes using static student personas, but relies on a human teacher in the loop, preventing automated optimisation. 

\section{Literature Synthesis Matrix}
A comparative summary of these seminal works is detailed in Table~\ref{tab:literature_synthesis}.

\begingroup
\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.3}
\begin{xltabular}{\linewidth}{>{\hsize=0.6\hsize}Y >{\hsize=1.3\hsize}Y >{\hsize=0.8\hsize}Y >{\hsize=1.15\hsize}Y >{\hsize=1.15\hsize}Y}
\caption{Comparative Literature Synthesis Matrix} \label{tab:literature_synthesis} \\
\toprule
\textbf{Paper \& Citation} & \textbf{Methodology / Core Achievement} & \textbf{Trajectory} & \textbf{Omission / Limitation} & \textbf{Our Niche \& Novelty} \\
\midrule
\endfirsthead

\multicolumn{5}{c}%
{{ Table \thetable:} Continued from previous page} \\
\toprule
\textbf{Paper \& Citation} & \textbf{Methodology / Core Achievement} & \textbf{Trajectory} & \textbf{Omission / Limitation} & \textbf{Our Niche \& Novelty} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{{Continued on next page}} \\
\endfoot

\bottomrule
\endlastfoot

Yuan et al. \cite{yuan2026towards} & Establishes Epistemic State Specification (ESS) for valid student simulation. & Formalising learner profiles in highly structured domains. & Assumes static learner capabilities and ignores cognitive fatigue. & Integrates dynamic Cognitive Bandwidth ($B_t$) equations. \\
\addlinespace
Scarlatos et al. \cite{scarlatos2026simulated} & Benchmarks simulated students; finds simple prompts are unreliable. & Supervised fine-tuning and DPO to calibrate profiles. & Focuses on evaluation; lacks active Socratic tutoring loops. & Builds cyclic Socratic loops with feedback guardrails. \\
\addlinespace
\"{O}zdemir \& Oztop \cite{ozdemir2026enhancing} & Tutor-student multi-agent interaction for code generation. & Peer-tutoring frameworks optimising code output. & Symmetric dialogues; lacks dynamic cognitive state tracking. & Models cognitive load ($B_t$) and friction ($F_t$) explicitly. \\
\addlinespace
Bian \cite{bian2026generative} & Interview-driven multi-expert reflection for learning behavioural simulation. & Using psychological domain experts to reflection-prompt. & Evaluates via static surveys (MSLQ); no active dialogue. & Deploys agents in dynamic LangGraph conversational loops. \\
\addlinespace
Liu et al. \cite{liu2024socraticlm} & Dean-Teacher-Student pipeline with SocraTeach dataset (35K dialogues). & Tuning models specifically for Socratic instruction. & Lacks safety check blocks; leaks answers easily. & Implements active Pydantic-enforced Guardrail Nodes. \\
\addlinespace
KELE \cite{peng2025kele} & Consultant-Teacher mechanism with SocRule (34 Socratic rules). & Structured, multi-stage rule-based Socratic teaching. & Science focus; lacks cognitive load models or sycophancy checks. & Targets programming education; models student bandwidth and scratchpad. \\
\addlinespace
Petukhova \& Kochmar \cite{petukhova2025intent} & Fine-grained taxonomy of 11 pedagogical intents. & Intent-annotated Socratic dialogue math training. & Math-specific; static intent mapping without active tracking. & Maps fine-grained Socratic actions ($\mathcal{A}$) dynamically via MDP. \\
\addlinespace
Xu et al. \cite{xu2025classroom} & Contextual student agents in virtual online classrooms. & Transferable Iterative Reflection (TIR) for courses. & Dyadic tutor optimisation is not explored. & Formulates tutoring strategy optimisation via MDP. \\
\addlinespace
GPTeach \cite{markel2023gpteach} & Simulated students with varied personas for TA training. & Safe sandbox environments for human teacher training. & No autonomous multi-agent tutor policy optimisation. & Automates Socratic tutoring policy optimisation (MDP). \\
\addlinespace
EduAgent \cite{xu2024eduagent} & Memory-augmented student agents to predict responses. & Structured memory logging across lessons. & Highly vulnerable to ``Yes-Man'' sycophancy bias. & Introduces Adversarial Scratchpad Probing checks. \\
\addlinespace
Maurya et al. \cite{maurya2025unifying} & 8-dimension unified taxonomy and MRBench benchmark. & Standardised assessment of LLM tutors on mathematics. & Focuses on evaluation; lacks active adaptive feedback loop. & Grounds evaluation annotation axes in structured Socratic categories. \\
\end{xltabular}
\endgroup

\section{Technical Notation and Key Theories}
To bridge the structural gaps in current literature, specifically the static assumption of learner profiles \cite{yuan2026towards} and the lack of mathematical models of student attention and cognitive friction in multi-agent dialogue \cite{ozdemir2026enhancing}, we establish a formal, mathematically rigorous foundation. We model Socratic instruction as a Partially Observable Markov Decision Process (POMDP), tracking hidden student belief states, modelling cognitive bandwidth constraints based on Cognitive Load Theory, and decoupling student compliance from genuine mastery using an information-theoretic bottleneck.

\subsection{POMDP Formulation of Socratic Dialogue}
The Socratic interaction is formulated as a Partially Observable Markov Decision Process (POMDP)\footnote{Modelling the dialogue as a POMDP is crucial because a student's internal cognitive state, such as their actual conceptual mastery and specific active misconceptions, is latent and cannot be directly read by the tutor. The tutor must infer this state solely from the student's natural language responses (observations).} defined by the tuple $(\mathcal{S}, \mathcal{A}, \mathcal{O}, \mathcal{T}, \mathcal{Z}, \mathcal{R}, \gamma)$ where:
\begin{itemize}
    \item \textbf{State Space ($\mathcal{S}$):} The student's hidden true cognitive state at turn $t$ is defined as:
    \begin{equation}
        s_t^{\text{true}} = \left(\mathbf{K}_t^{\text{true}}, \mathbf{M}_t^{\text{true}}, B_t, F_t\right)
    \end{equation}\footnote{By including continuous cognitive variables ($B_t, F_t$) alongside discrete mastery and misconception states ($\mathbf{K}_t^{\text{true}}, \mathbf{M}_t^{\text{true}}$), the state space captures both the long-term knowledge structure and short-term working memory dynamics of the learner. Continuous variables capture immediate cognitive strain, while discrete vectors track cognitive schemas.}, where:
    \begin{itemize}
        \item $\mathbf{K}_t^{\text{true}} \in [0, 1]^m$ is the student's true latent mastery vector across $m$ core programming concepts, structured as a Directed Acyclic Graph (DAG) $\mathcal{G}_{\text{prereq}} = (\mathcal{V}, \mathcal{E})$ representing prerequisite dependencies.\footnote{Each element $K^{\text{true}}_{i, t}$ is continuous to represent degrees of conceptual mastery, where $0$ denotes complete lack of familiarity, and $1$ represents fluent independent recall and execution capability. The DAG edges $(j, i) \in \mathcal{E}$ define prerequisite constraints, such that mastery of parent concepts (e.g., control loops) is a gating condition for learning child concepts (e.g., nested iteration), preventing ungrounded mastery transitions.}
        \item $\mathbf{M}_t^{\text{true}} \in \{0, 1\}^k$ represents the presence or absence of $k$ active conceptual misconceptions.\footnote{Unlike mastery scores which are continuous, misconceptions are modelled here as discrete boolean flags ($\mathbf{M}_t^{\text{true}} \in \{0, 1\}^k$), representing specific faulty mental models (e.g., confusing assignment with cloning) that are either active or inactive. This supports targeted Socratic counterexamples to destabilise specific misconceptions.}
        \item $B_t \in [0, 1]$ represents the student's available Cognitive Bandwidth (working memory capacity).
        \item $F_t \in [0, 1]$ represents active Cognitive Friction (instructional alignment mismatch).
    \end{itemize}
    Unlike monolithic dialogue architectures that rely on ephemeral history logs, we maintain this high-dimensional state space using a centralised, versioned learner state buffer \cite{intellicode2025}. To resolve concurrency conflicts and prevent state synchronisation race conditions in multi-agent executions, we enforce a strict single-writer policy where only the Epistemic Tracker is authorised to commit write operations to the buffer, while the Socratic Tutor and Simulated Student nodes interface with state snapshots as pure, read-only functions \cite{elhaimeur2026itas}.
    \item \textbf{Action Space ($\mathcal{A}$):} The set of Socratic pedagogical directives available to the tutor agent (e.g., \textit{deconstruct\_code}, \textit{present\_analogy}, \textit{prompt\_prediction}, \textit{ask\_counterfactual}, \textit{explain\_concept}). This structured action space aligns with the fine-grained pedagogical intent taxonomy proposed by Petukhova \& Kochmar \cite{petukhova2025intent}, where setting specific pedagogical intents rather than coarse directives is shown to significantly improve response alignment and pedagogical effectiveness.
    \item \textbf{Observation Space ($\mathcal{O}$):} The space of natural language dialogue responses generated by the student, denoted $o_t$.
    \item \textbf{Transition Function ($\mathcal{T}$):} The probability distribution of transitioning to a new state $s_{t+1}$ given the current state $s_t$ and tutor action $a_t$:
    \begin{equation}
        \mathcal{T}(s_{t+1} \mid s_t, a_t) = P(s_{t+1} \mid s_t, a_t)
    \end{equation}
    which captures how the student's knowledge vector, misconceptions, bandwidth, and friction mutate in response to the tutor's pedagogical choices.\footnote{We note a temporal asymmetry in the state component transitions: while the knowledge vector $\mathbf{K}_{t+1}$ transitions as a consequence of the learning experience from the tutor's prompt $a_t$ at turn $t$, the cognitive variables $B_{t+1}$ and $F_{t+1}$ update immediately upon exposure to the next pedagogical prompt $a_{t+1}$ at turn $t+1$ to reflect the immediate cognitive load of the incoming explanation. This decoupling reflects the difference between long-term memory schema consolidation (slow dynamics) and working memory capacity (fast dynamics).}
    \item \textbf{Observation Function ($\mathcal{Z}$):} The probability of observing dialogue response $o_t$ given the transitioned state $s_t^{\text{true}}$ and the tutor's current action prompt $T_t$:
    \begin{equation}
        \mathcal{Z}(o_t \mid s_t^{\text{true}}, T_t) = P(o_t \mid s_t^{\text{true}}, T_t)
    \end{equation}
    which is parameterised directly by the student's generator response policy $\pi_{\text{student}}(o_t \mid T_t, \mathbf{K}_t^{\text{true}}, \mathbf{M}_t^{\text{true}}, B_t, F_t)$.\footnote{The observation probability conditions on the transitioned state $s_t^{\text{true}}$ and current action prompt $T_t$. In the dialogue setting, the student response $o_t$ is generated after receiving the tutor's prompt $T_t$, reflecting both the prompt structure and the student's new cognitive state $s_t^{\text{true}}$ resulting from that prompt.}
    \item \textbf{Reward Function ($\mathcal{R}$):} The feedback signal designed to optimise pedagogical efficiency, penalise tutor direct-answer leakage, and avoid cognitive overload:
    \begin{equation}
        R_t = \sum_{i=1}^m w_i \left(K^{\text{true}}_{i, t+1} - K^{\text{true}}_{i, t}\right) - w_f F_t - w_b (1 - B_t) - w_w W_t
    \end{equation}
    where $w_i > 0$ are concept importance weights, $w_f > 0$ is the cognitive friction penalty, $w_b > 0$ is the working memory overload penalty, and $w_w > 0$ is a severe penalty triggered if the Guardrail Node detects direct answer or code leakage ($W_t \in \{0, 1\}$). Because the student's true latent state is hidden from the tutor, the Socratic tutor policy is optimised to maximise the \textbf{expected reward} under the tracker's belief state $b_t(s)$\footnote{The expected reward function balances the primary educational goal (concept learning gain $\hat{K}_{i, t+1} - \hat{K}_{i, t}$) against secondary constraints (friction, bandwidth overload, and safety guardrails). In policy optimisation, this expected reward is used to train or prompt the tutor policy to discover the optimal Socratic trajectory, maximising cumulative expected discounted reward $\mathbb{E}[\sum_t \gamma^t \bar{R}_t]$}:
    \begin{equation}
        \bar{R}_t = \mathbb{E}_{s \sim b_t}[R_t(s, a_t)] = \sum_{i=1}^m w_i \left(\hat{K}_{i, t+1} - \hat{K}_{i, t}\right) - w_f \hat{F}_t - w_b (1 - \hat{B}_t) - w_w W_t
    \end{equation}
    where $\hat{K}_{i, t}$ is the expected concept mastery, and $\hat{B}_t, \hat{F}_t$ are the expected cognitive bandwidth and cognitive friction estimated by the tracker under the belief distribution.
    
    The Socratic tutor policy $\pi(a_t \mid s'_t)$ is defined over a discretised representation of the expected cognitive state space $s'_t \in \mathcal{S}_{\text{discrete}}$ (which maps continuous estimates $\hat{\mathbf{K}}_t$ and active misconceptions $\hat{\mathbf{M}}_t$ to student archetypes). To guarantee computational and financial feasibility within the 14-week project timeline, avoiding the high token cost and slow convergence of online deep reinforcement learning, we optimise prompt-selection using Q-learning over this discrete state space. The tutor learns the action-value function $Q(s'_t, a_t)$ mapping states to discrete pedagogical prompt templates $a_t \in \mathcal{A}$ via the temporal-difference learning recurrence:
    \begin{equation}
        Q(s'_t, a_t) \gets (1 - \alpha_{\text{QL}}) Q(s'_t, a_t) + \alpha_{\text{QL}} \left( \bar{R}_t + \gamma \max_{a' \in \mathcal{A}} Q(s'_{t+1}, a') \right)
    \end{equation}
    where $\alpha_{\text{QL}} \in (0, 1)$ is the learning rate, and the Socratic policy selects prompts greedily: $\pi(s'_t) = \arg\max_{a \in \mathcal{A}} Q(s'_t, a)$.
    \item \textbf{Discount Factor ($\gamma \in [0, 1]$):} Weighs immediate learning gains against long-term instructional sustainability.
\end{itemize}


